---
title: "Final Project: Geospatial Regressions"
subtitle: "Data Science for Public Policy"
authors: "Akbar Syed Naqvi, Thomas Tiberghien Alvarez"
format: 
  html:
    code-line-numbers: true
editor: visual
execute: 
  warning: false
self-contained: true
urlcolor: blue
---

## Background

Regressions are considered the bread and butter of data analysis. They have been long used in economics and social science to establish causal relationships, but data science has utilized regression methods as a backbone for prediction. Spatial regressions take the power of regressions and add an element of spatial dependency.

Space can impact how one generates data and be a form of systematic error in our analysis. Spatial analysis in its modern form has been implemented and iterated upon through many subjects: biology through plant distributions and animal movement, epidemiology through disease mapping, and most notably economics through spatial econometrics, Spatial structure helps regression model in two ways

1.  Process of generating data is spatial e.g. price of homes often dependent on area

2.  Geography can allow you to assess structure of errors and mispredictions if they are systematic, often times errors are assumed to be independently and identically distributed, but this assumption falls apart with spatial data

Spatial regression is also easier to visualize than just rows and columns of data. This makes it easier to explain insights more broadly, rather than trying to tell a story through a few data points. 

## Methods

Source: Geographic Data Science in Python

Regression models can sometimes display clustering in the error terms, which is a potential violation of i.i.d assumption in linear models. To interrogate whether or not there is clustering, one can observe distribution of error terms dis-aggregated by a spatial element of concern, visualizes how the median income in a state's counties differs by their distance to a major metro area, for example. An example of a of more sophisticated test that can be used to decipher a spatial element is a spatial autocorrelation test. Often times in non-spatial regressions, one can use spatial autocorrelation tests like Moran's I, Geary's c, or Getis and Ord's G-statistic on residuals. If spatial autocorrelations exist, this can justify the use of spatial model


    -   Often times in non-spatial regressions, one can use spatial autocorrelation tests like Moran's I, Geary's c, or Getis and Ord's G-statistic on residuals

**Spatial Weights**


Spatial questions target specific information about the spatial configuration of an outcome variable. When it comes to spatial analysis, we are often building a topography, a mathematical structure that expresses connectivity between observations. This weighting can help us create a *geographically weighted regression*; local version of spatial regression that generates parameters dis aggregated by spatial units of analysis , partial weights are one way to express this topology. Examples of spatial weights include

-   Contiguity Weights - A contiguous pair of spatial objects who share a common border,

-   Distance based weights - Proximity based, often observing the nearest neighbors, requires defining k nearest neighbors with spatial information used to calculate centroids

-   Block Weights - We weigh spatial elements based on a specific geographic group they belong to, and any elements not in the group disconnected from those observations, e.g. connecting counties to states

**Bringing Space Into a Regression**

Spatial regression involves introducing geographic space or context into our regressions when we feel that it plays a role or can act as a proxy for other unobserved variables. There are a variety of methods one can use to do this.

-   Spatial Feature Engineering

    -   In data science, feature engineering involves applying domain knowledge to raw data in order to structure it in a way that is meaningful, in other words, transforming a dataset to prepare it for analysis

    -   Geography is one of the best ways to introduce domiain knowledge into a data science problem.

    -   Spatial feature engineering is the process of developing additional information from raw data using geographical knowledge

    -   Simplest example -

## Application

Source: https://walker-data.com/census-r/modeling-us-census-data.html

Now that we've discussed the background and various methods used in spatial regressions, let's utilize the Census's American Community Survey (ACS) data to see how it all plays out in practice. Demographic statistics are common covariates utilized in linear regression, but they run into issues with spatial autocorrelation, which makes ACS data a great candidate for spatial regression techniques.

First, we define a set of counties we want to analyze, and return a set of variables, using the *tidycensus* package. We load in the sf package as well so we can do a spatial join later.

```{r}

library(units)
library(sf)
library(tidycensus)

dfw_counties <- c("Collin County", "Dallas", "Denton", 
                  "Ellis", "Hunt", "Kaufman", "Rockwall", 
                  "Johnson", "Parker", "Tarrant", "Wise")

variables_to_get <- c(
  median_value = "B25077_001",
  median_rooms = "B25018_001",
  median_income = "DP03_0062",
  total_population = "B01003_001",
  median_age = "B01002_001",
  pct_college = "DP02_0068P",
  pct_foreign_born = "DP02_0094P",
  pct_white = "DP05_0077P",
  median_year_built = "B25037_001",
  percent_ooh = "DP04_0046P"
)

dfw_data <- get_acs(
  geography = "tract",
  variables = variables_to_get,
  state = "TX",
  county = dfw_counties,
  geometry = TRUE,
  output = "wide",
  year = 2020
) %>%
  select(-NAME) %>%
  st_transform(32138) # NAD83 / Texas North Central

```

Next, we do some feature engineering by adding population density which represents the number of people in each Census tract per square kilo and median_structure_age, which represents median age of housing structures in tract.

```{r}

dfw_data_for_model <- dfw_data %>%
  mutate(pop_density = as.numeric(set_units(total_populationE / st_area(.), "1/km2")),
         median_structure_age = 2018 - median_year_builtE) %>%
  select(!ends_with("M")) %>% 
  rename_with(.fn = ~str_remove(.x, "E$")) %>%
  na.omit()
```

Like indicated before, the main idea behind spatial autocorrelation is that the independence of residuals are violated because the model's performance depends on geographic location.

One technique we discussed before for testing spatial autocorrelation was Moran's *I.* Let's use it

```{r}

library(spdep)

wts <- dfw_data_for_model %>%
  poly2nb() %>%
  nb2listw()

moran.test(dfw_data_for_model$residuals)

dfw_data_for_model$lagged_residuals <- lag.listw(wts, dfw_data_for_model$residuals)

ggplot(dfw_data_for_model, aes(x = residuals, y = lagged_residuals)) + 
  theme_minimal() + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", color = "red")
```

If we observe that there is spatial autocorrelation, we can use a spatial error model in order to account for this, where we capture latent spatial processes not accounted for in the model's estimation being captured in the residuals. This can be done with the errorsarlm() function in spatialreg.

```{r}

error_model <- errorsarlm(
  formula = formula2, 
  data = dfw_data_for_model, 
  listw = wts
)

summary(error_model, Nagelkerke = TRUE)
```

We look at lambda to confirm if it is large and statistically significant, which confirms to us the importance of accounting for spatial autocorrelation.
